A Simple Type Theory for Category Theory

### 2022-05-27 paper/paper.tex

\begin{definition}
  We define composition of transformation contexts $\Phi \jnctx \Psi$ by induction on $\Psi$
  \begin{align*}
    \Phi \jnctx \beta:\cat D &= \Phi    \\
    \Phi \jnctx \Psi,x:R,\beta:\cat D &= (\Phi\jnctx\Psi),x:R,\beta:\cat D
  \end{align*}
\end{definition}

It follows by an inductive argument that this definition is
associative and unital.

\begin{definition}
  We define the horizontal composition of transformation substitutions
  $\phi \jnctx \psi$ whenever $d^+\phi = d^-\psi$ by induction on
  $\psi$:
  \begin{align*}
    \phi \jnctx a/\alpha &= \phi\\
    \phi \jnctx (\psi,t/x,a/\alpha) &= (\phi \jnctx \psi),t/x,a/\alpha
  \end{align*}
\end{definition}

It follows by an inductive argument that this operation is associative
and unital.

\begin{lemma}[Inversion]
  \begin{enumerate}
  \item If $\Phi \vdash \psi :: (\alpha:\cat C)$ then $\Phi =
    \beta:\cat D$ for some $\cat D$ and $\psi = a/\alpha$ where
    $\beta:\cat D \vdash a : \cat C$.
  \item If $\Phi \vdash \psi :: \Psi_1 \jnctx \Psi_2$, then there
    exists unique $\Phi_1,\Phi_2,\psi_1,\psi_2$ such that $\Phi =
    \Phi_1\jnctx\Phi_2$ and $\Phi_1\vdash\psi_1 :: \Psi_1$ and
    $\Phi_2\vdash \psi_2 : \Psi_2$ and $\psi = \psi_1\jnctx\psi_2$.
  \end{enumerate}
\end{lemma}

\begin{definition}
  We define the vertical composition of transformation substitutions
  $\phi[\psi]$ inductively on $\phi$.
  \begin{align*}
    (a/\alpha)[b/\beta] &= a[b/\beta]/\alpha\\
    (\phi_1,t/x,a/\alpha)[\psi_1\jnctx\psi_2] &= \phi_1[\psi_1], t[\psi_2],a[d^+\psi_2]
  \end{align*}
  This covers all cases by \cref{inversion}.

  We define the vertical identity $\id_\Phi$ by induction on $\Phi$
  \begin{align*}
    \id_{\alpha:\cat C} &= \alpha/\alpha\\
    \id_{\Phi,x:R,\alpha:\cat C} &= \id_{\Phi},x/x,\alpha/\alpha
  \end{align*}
\end{definition}
By an inductive argument, this operation is associative and unital.

\begin{lemma}{Interchange}
  We prove the interchange law
  $$(\phi\jnctx\psi)[\phi'\jnctx\psi'] = \phi[\phi'] \jnctx \psi[\psi']$$
  by induction on $\psi$.

>>>>

In this section we establish the most significant portion of the
completeness proof:
\begin{construction}[Double category completeness]
  For any $\Gamma$, define a strict double category as follows
  \begin{enumerate}
  \item Objects are categories $\Gamma \vdash \cat C \isaCat$
  \item Vertical morphisms from $\cat C$ to $\cat D$ are functor/open
    object terms $\Gamma \pipe\alpha:\cat C \vdash B : \cat D$ modulo
    $\alpha$-equivalence.
  \item Horizontal morphisms from $\cat C$ to $\cat D$ are profunctor
    contexts $\Gamma\vdash \Phi \isavectx$ with $d^-\Phi = \cat C$ and
    $d^+\Phi = \cat D$ modulo $\alpha$-equivalence.
  \item Squares from $\Psi$ to $\Phi$ along $A$ and $B$ are
    substitutions $\Gamma \vdash \phi : \Psi \to \Phi$ with $d^-\phi =
    A$ and $d^+\phi = B$ modulo $\alpha$-equivalence.
  \end{enumerate}
\end{construction}
\begin{proof}
  The details are in the linked lemmas. Here we give an overview of
  the relevant structure and properties.
  \begin{enumerate}
  \item Vertical composition is given by substitution of object terms
    and is associative and unital with the variable as the identity
    (Lemma \ref{lem:vertical-cat}).
  \item Horizontal composition is given by the gluing of contexts
    $\Phi\jnctx \Psi$ and is associative and unital with the single
    category variable as the identity (Lemma
    \ref{lem:horizontal-cat}).
  \item Vertical composition of substitutions is given by substitution
    $\phi[\psi]$ and is associative and unital with the identity
    substitution given by a substitution of variables for themselves
    (Lemma \ref{lem:vertical-sq-cat}).
  \item Horizontal composition of substitutions is given by gluing of
    substitutions $\phi \jnctx \psi$ and is associative and unital
    with the identity substitution given by a single functor/open
    object substitution (Lemma \ref{lem:horizontal-sq-cat}).
  \item The process of constructing horizontal identity substitutions
    is functorial in that the horizontal identity of a vertical
    identity arrow is the vertical identity
    (Lemma~\ref{lem:hor-vert-id-id}) and the vertical composition of
    horizontal identities is a horizontal identity
    (Lemma~\ref{lemma:lem-hor-sq-id}).
  \item The process of horizontal composition of substitutions is
    functorial in that the horizontal composition of vertical identity
    squares is the identity (Lemma~\ref{lem:lem-vert-sq-id}) and the
    interchange law holds (Lemma~\ref{lem:interchange}).
  \end{enumerate}
\end{proof}

\begin{lemma}[Vertical category]
  \label{lem:vertical-cat}
  Substitution of object terms $A[B/\beta]$ is associative with
  identity given by the variable $\alpha:\cat C \vdash \alpha:\cat C$.
\end{lemma}

\begin{definition}
  \label{lem:horizontal-cat}
  We define composition of transformation contexts $\Phi \jnctx \Psi$ by induction on $\Psi$
  \begin{align*}
    \Phi \jnctx \beta:\cat D &= \Phi    \\
    \Phi \jnctx \Psi,x:R,\beta:\cat D &= (\Phi\jnctx\Psi),x:R,\beta:\cat D
  \end{align*}
  By induction, this is seen to be associative
  \[ (\Phi \jnctx \Psi) \jnctx \Sigma = \Phi \jnctx (\Psi \jnctx \Sigma) \]
  and unital with identity for $\cat C$ given by the single category
  variable context $\alpha:\cat C$
  \[ \alpha:\cat C \jnctx \Phi = \Phi = \Phi \jnctx \beta:\cat D\]
  when $d^-\Phi = \alpha:\cat C$ and $d^+\Phi = \beta:\cat D$.
\end{definition}

\begin{definition}
  \label{lem:horizontal-sq-cat}
  We define the horizontal composition of transformation substitutions
  $\phi \jnctx \psi$ whenever $d^+\phi = d^-\psi$ by induction on
  $\psi$:
  \begin{align*}
    \phi \jnctx a/\alpha &= \phi\\
    \phi \jnctx (\psi,t/x,a/\alpha) &= (\phi \jnctx \psi),t/x,a/\alpha
  \end{align*}
  By induction, this is seen to be associative
  \[ \phi \jnctx (\psi \jnctx \sigma) = (\phi \jnctx \psi) \jnctx \sigma \]
  and unital with identity given by a single object substitution
  $a/\alpha : \alpha:\cat C \to \beta:\cat D$
  \[ a/\alpha \jnctx \phi = \phi = \phi \jnctx b/\beta \]
  when $d^-\phi = a/\alpha$ and $d^+\phi = b/\beta$.
\end{definition}

To define vertical composition of squares, we need the following lemma.
\begin{lemma}[Inversion]
  \begin{enumerate}
  \item If $\Phi \vdash \psi :: (\alpha:\cat C)$ then $\Phi =
    \beta:\cat D$ for some $\cat D$ and $\psi = a/\alpha$ where
    $\beta:\cat D \vdash a : \cat C$.
  \item If $\Phi \vdash \psi :: \Psi_1 \jnctx \Psi_2$, then there
    exists unique $\Phi_1,\Phi_2,\psi_1,\psi_2$ such that $\Phi =
    \Phi_1\jnctx\Phi_2$ and $\Phi_1\vdash\psi_1 :: \Psi_1$ and
    $\Phi_2\vdash \psi_2 : \Psi_2$ and $\psi = \psi_1\jnctx\psi_2$.
  \end{enumerate}
\end{lemma}

\begin{definition}
  \label{lem:vertical-sq-cat}
  We define the vertical composition of transformation substitutions
  $\phi[\psi]$ inductively on $\phi$.
  \begin{align*}
    (a/\alpha)[b/\beta] &= a[b/\beta]/\alpha\\
    (\phi_1,t/x,a/\alpha)[\psi_1\jnctx\psi_2] &= \phi_1[\psi_1], t[\psi_2],a[d^+\psi_2]
  \end{align*}
  This covers all cases by \cref{inversion}.

  We define the vertical identity $\id_\Phi$ by induction on $\Phi$
  \begin{align*}
    \id_{\alpha:\cat C} &= \alpha/\alpha\\
    \id_{\Phi,x:R,\alpha:\cat C} &= \id_{\Phi},x/x,\alpha/\alpha
  \end{align*}

  By induction this is seen to be associative:
  \[ \phi[\psi][\sigma] = \phi[\psi[\sigma]] \]
  and unital
  \[ \id_{\Phi}[\phi] = \phi = \phi[\id_{\Psi}] \]
\end{definition}

\begin{lemma}
  \label{lem:sq-vid-hid}
  For any $\cat C$, $\alpha/\alpha : \alpha:\cat C \to \alpha:\cat C$
  is both a horizontal and vertical identity.
\end{lemma}

\begin{lemma}
  \label{lem:sq-vcomp-hid}
  For any $\beta:\cat D \vdash a : \cat C$ and $\alpha:\cat C \vdash a': \cat C'$
  the substitution
  \[ (a'/\alpha')[a/\alpha] : \beta:\cat D \to \alpha':\cat C' \]
  satisfies
  \[ (a'/\alpha')[a/\alpha] = a'[a/\alpha]/\alpha' \]
\end{lemma}

\begin{lemma}
  \label{lem:sq-vid-hcomp}
  For any $\Phi$ and $\Psi$ with $\Phi\jnctx \Psi$ well-defined,
  \[ \id_\Phi \jnctx \id_\Psi = \id_{\Phi\jnctx\Psi} \]
  follows by induction.
\end{lemma}

\begin{lemma}{Interchange}
  \label{lem:sq-vcomp-hcomp}
  We prove the interchange law
  $$(\phi\jnctx\psi)[\phi'\jnctx\psi'] = \phi[\phi'] \jnctx \psi[\psi']$$
  by induction on $\psi$.
